Title Text

### 2013-05-06 papers/gpce2013/gpce2013.tex
intro

\title{Title Text}
\subtitle{Subtitle Text, if any}

@@

Rich Internet Applications have interesting properties: they require no installation or deployment on
clients and enables large scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. For instance, HTML fragments may be
built from the server-side when a page is requested by a client, but they may also be built from the client-side to
perform an incremental update subsequent to a user action. How could developers write HTML fragment definitions once
and render them on both client-side and server-side? The more interactive the application is, the more logic needs to
be duplicated between the server-side and the client-side.

Using the same programming language on both server-side and client-side could reduce the gap between client-side and
server-side environments, however the JavaScript language -- which is currently the only action language natively
supported by all Web browsers -- has several drawbacks making it hardly suitable for large code bases (\eg no static
typing, no module system, verbose syntax, \etc). An increasing number of programming languages support generating
JavaScript (\eg GWT, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}), increasing the panel
of programming languages available to write both the client-side and the server-side code of a Web application.

Having a full featured, cutting edge, programming language that runs on both client-side and server-side can help
developers to write more maintainable code, however by abstracting over the differences of the client-side and the
server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
applications, because they are supposed to run on a broad range of devices, from the powerful desktop personal
computer to the less powerful smartphone. “Every 100~ms delay costs 1\% of sales”, said Amazon in 2006.

For instance, the boundaries of the code parts to emit on client-side are less visible when you share code between
client-side and server-side so transitive dependencies may pull a lot of code, causing a high download overhead.
Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way. (TODO Give a
concrete example for each problem)

On one hand, for engineering reasons, developers want to write Web applications using a single programming language,
abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
control on the way their code is compiled to each target platform. How to solve this dilemma?

On top of this previous work, this paper presents several new APIs (?) that can be shared between client-side and
server-side code and/or that are efficiently translated on each target platform. More precisely, our contributions
are:

\begin{itemize}
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Use monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side ;
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs
\end{itemize}

@@

\section{Related Works}

@@

\section{Implementation}

>>>>

\title{Title Text}
\subtitle{You don’t have to trade abstraction for control}

@@

Web applications are attractive because they require no installation or deployment on clients and enable large scale
collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintainance effort: once you decided to implement a feature on
client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
Even worse, logic parts that need to run on both client-side and server-side are duplicated. For instance HTML
fragments may be built from the server-side when a page is requested by a client, but they may also be built from the
client-side to perform an incremental update subsequent to an user action. How could developers write HTML fragment
definitions once and render them on both client-side and server-side? The more interactive the application is, the
more logic needs to be duplicated between the server-side and the client-side (explain why?).

Using the same programming language on both server-side and client-side can improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
only action language natively supported by almost all Web clients -- can be used on server-side, and an increasing
number of programming languages or compiler backends can generate JavaScript code (\eg
Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

However, this engineering comfort may come at the price of an inefficient runtime: abstracting over platform
differences often means restricting to a subset of common features and losing opportunities to perform platform
specific optimizations. Performance is a primary concern in Web applications, because they are expected to run on a
broad range of devices, from the powerful desktop personal computer to the less powerful smartphone. “Every 100~ms
delay costs 1\% of sales”, said Amazon in 2006. For instance, because the boundaries of the code sent to the client
are less visible when you share code between client-side and server-side, transitive dependencies may pull a lot of
code on the client, causing a high download overhead. Moreover, generating efficient code for heterogeneous platforms
is hard to achieve in an extensible way: the translation of common abstractions like collections into their native
counterpart (JavaScript arrays on client-side and standard library’s collections on server-side) may be hardcoded in
the compiler, but that would not scale to handle all the abstractions a complete application may use (\eg HTML
fragment definitions, form validation rules, or even some business data type that may be represented differently for
performance reasons).

On one hand, for engineering reasons, developers want to write Web applications using a single language, abstracting
over the target platforms differences. But on the other hand, for performance reasons, they want to keep
control on the way their code is compiled to each target platform. How to solve this dilemma?

Compiled domain specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain specific
languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target platform.

This paper presents such a DSL allowing developers to write Web applications in a single language which code
fragments can be shared between client and server sides, and which is efficiently compiled to each side. More
precisely, we demonstrate the following features:

\begin{itemize}
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs.
\end{itemize}

@@

\section{Related Work}

@@

\section{Implementation?}
